Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility

### 2025-12-03 paper/paper.tex
Up draft presentation

  \author{Andreas Laffert Tamayo}
            \affil{%
                  Instituto de Sociología, Pontificia Universidad
                  Católica de Chile
              }

\date{}

% Colores extra por nombre
\usepackage[dvipsnames]{xcolor}
% Ajustes de hipervínculos
% \hypersetup{
%   colorlinks = true,
%   citecolor  = DarkSlateBlue,     % << citas
%   urlcolor   = DarkSlateBlue  % URLs/DOIs
% }
% 
\begin{document}
\maketitle
\begin{abstract}
My abstract \newline \textbf{Keywords}: Pension commodification ·
Intergenerational mobility · Causal inference · Meritocracy · Chile
\end{abstract}

\setstretch{1.5}
This document was last modified at 2025-12-01 02:26:24 and it was last
rendered at 2025-12-01 02:26:24.

\section{Introduction}\label{introduction}

@@

Social mobility and meritocratic perceptions are closely intertwined in
the acculturation mechanism, although the direction of this link can
vary across trajectories and contexts. A large body of work consistently
shows that upwardly mobile individuals tend to interpret their
trajectory through a meritocratic lens, reading their success as
evidence that effort and ability are rewarded and reinforcing
individualistic attributions of poverty and wealth
(\citeproc{ref-bucca_merit_2016}{Bucca, 2016};
\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b};
\citeproc{ref-mijs_belief_2022}{Mijs et al., 2022};
\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}), even in a
highly unequal and low mobility setting were the narratives of
individual success reinforces (\citeproc{ref-deng_its_2025}{Deng \&
Wang, 2025}). Studies of economic elites in Chile similarly report
strong endorsement of meritocracy, with respondents attributing their
ascent mainly to talent and business or leadership skills
(\citeproc{ref-atria_economic_2020}{Atria et al., 2020}). Evidence on
downward or blocked mobility is more ambivalent. Exposure to
low-mobility contexts and experiences of downward mobility are
associated with weaker beliefs that effort is rewarded, lower just-world
and opportunity beliefs, and stronger structural attributions for
success and failure (\citeproc{ref-davidai_why_2018}{Davidai, 2018};
\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b};
\citeproc{ref-mijs_how_2022}{Mijs \& Hoy, 2022}). Yet, under strong
meritocratic convictions, status loss can also be internalised as
personal failure, sustaining system-justifying attitudes and the
legitimation of existing hierarchies as a psychological defence strategy
(\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
\citeproc{ref-deng_its_2025}{Deng \& Wang, 2025};
\citeproc{ref-mau_inequality_2015}{Mau, 2015}). In sum, upward mobility
is generally associated with stronger meritocratic perceptions, whereas
downward mobility is more closely linked to structural explanations of
inequality, without necessarily eroding meritocratic convictions.

The ways in which social mobility trajectories shape interpretations of
meritocracy can determine whether mobility translates into stronger or
weaker support for market-based pension welfare. Following Mau's
(\citeproc{ref-mau_inequality_2015}{2015}) account of the ``majority
class'', upward mobility is expected to strengthen market-based welfare
not only because material interests change, but because those who rise
come to read their story as one of individual effort and talent
rewarded. In this view, markets are seen as fair arenas that properly
allocate pensions and other welfare benefits according to merit, so
upwardly mobile individuals with strong meritocratic convictions should
be especially prone to endorse commodified pensions. In contrast,
downward mobility typically fosters de-commodifying demands, yet under
strong meritocratic perceptions, status loss may be read as individually
deserved, weakening---or even reversing---its usual de-commodifying
pattern, as market outcomes are still viewed as broadly fair regardless
of direction. Empirical research on redistribution and attitudes toward
inequality supports this mechanism. Survey studies suggest that
subjective social mobility shapes support for redistribution and pension
spending through self-serving bias
(\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}), although recent
work indicates that this mechanism is evident only under explicit
constraints in experimental designs
(\citeproc{ref-molina_its_2019}{Molina et al., 2019}). More in line with
the acculturation theories, experimental reserach has shown that
exposure to upward-mobility information or self-made narratives reduces
support for redistribution, increases tolerance of inequality and system
justification, and even legitimises exploitative work conditions,
whereas exposure to low or downward mobility produces the opposite
pattern; crucially, these effects are partly mediated by stronger
meritocratic perceptions (\citeproc{ref-deng_its_2025}{Deng \& Wang,
2025}; \citeproc{ref-matamoros-lima_social_2025}{Matamoros-Lima et al.,
2025}; \citeproc{ref-shariff_income_2016}{Shariff et al., 2016}). In
this framework, meritocratic perceptions condition the attitudinal
consequences of mobility by amplifying the pro-market effects of upward
mobility and muting the critical potential of downward mobility; thus, I
treat meritocratic perceptions as a moderating mechanism linking
intergenerational mobility to preferences for pension commodification.

@@

The findings reveal a tension in Chileans' judgements of market-based
pension welfare. Across 2016, 2018 and 2023, large majorities reject the
claim that it is fair for higher-income people to obtain better pensions
than those with lower incomes. However, a non-trivial---and
growing---minority endorses this principle, with the sharpest increase
between 2018 and 2023. Market-based pension allocation thus remains a
minority view. However, it is becoming more common, in line with
longitudinal evidence of rising support for income-based access to
welfare services in Chile
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}). This coexistence unfolds in a system of mandatory, privately
managed, defined-contribution accounts that tightly link old-age income
to earnings, contribution density and financial returns. This design has
produced low and unequal pensions and state interventions against
old-age poverty (\citeproc{ref-galvez_pensiones_2024}{Gálvez et al.,
2024}; \citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al.,
2019}), while framing pensions as personal investments rather than
solidarity-based rights and promoting individual responsibility and
contribution-based desert
(\citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012}). The Chilean
case fits evidence that supports the claim that income-based welfare
access is stronger in liberal, marketised regimes
(\citeproc{ref-lindh_public_2015}{Lindh, 2015}). This coexistence also
echoes recent pension conflict---from the ``No+AFP'' protests to
post-pandemic withdrawals and the ``Con mi plata NO'' mobilisation
(\citeproc{ref-galvez_con_2023}{Gálvez, 2023})---and illustrates mixed
policy feedback: the same regime generates rights-based critiques and
proprietarian defences of pension commodification. Together, these
patterns show that, despite discontent and a legitimacy crisis, support
for pension commodification remains salient, raising the question of
which social trajectories and beliefs underpin it.

@@

The second set of hypotheses examined whether meritocratic perceptions
condition the effects of intergenerational mobility. Building on
acculturation accounts and self-serving bias arguments
(\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}), I expected mobility to affect preferences for
pension commodification not only through material self-interest but also
through shifts in normative frames of distributive justice. Prior
research shows that upwardly mobile individuals tend to interpret their
trajectory through a meritocratic lens, reading their success as
evidence that effort and ability are rewarded and reinforcing
individualistic attributions of poverty and wealth. In contrast,
downward or blocked mobility is more often associated with structural
explanations of inequality and more ambivalent meritocratic beliefs,
sometimes including the internalisation of status loss as personal
failure (\citeproc{ref-atria_economic_2020}{Atria et al., 2020};
\citeproc{ref-bucca_merit_2016}{Bucca, 2016};
\citeproc{ref-davidai_why_2018}{Davidai, 2018};
\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
\citeproc{ref-deng_its_2025}{Deng \& Wang, 2025};
\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b};
\citeproc{ref-mijs_how_2022}{Mijs \& Hoy, 2022}). Following Mau's
(\citeproc{ref-mau_inequality_2015}{2015}) account of the ``majority
class'', I therefore expected upward mobility under strong meritocratic
convictions to be especially conducive to endorsing commodified
pensions, and strong meritocracy to dampen the usual de-commodifying
effect of downward mobility. Empirically, however, the interaction
models offer little support for this mechanism. Meritocratic perceptions
do not systematically moderate the mobility effects identified in the
primary analysis, and the few trajectory-specific patterns that
tentatively point in the expected direction are small and imprecisely
estimated. This result contrasts with survey and experimental studies
showing that upward mobility experiences or exposure to self-made
narratives reduce support for redistribution and public pension spending
partly by reinforcing meritocratic convictions
(\citeproc{ref-deng_its_2025}{Deng \& Wang, 2025};
\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
\citeproc{ref-matamoros-lima_social_2025}{Matamoros-Lima et al., 2025};
\citeproc{ref-molina_its_2019}{Molina et al., 2019};
\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011};
\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}). One
possibility is that, in this context, intergenerational occupational
mobility---as operationalised here---has limited impact on meritocratic
perceptions, which appear more weakly tied to mobility trajectories than
previous research suggests and more strongly anchored in broader
socialisation processes and class-specific cultures.

This study makes several contributions. Substantively, it offers one of
the first systematic analyses of preferences for pension commodification
in Chile's highly unequal and marketised pension regime, addressing
Mau's (\citeproc{ref-mau_inequality_2015}{2015}) question of why core
elements of neoliberal welfare arrangements gain public acceptance. By
showing that upward and downward intergenerational mobility into and out
of upper-class positions causally shape support for market-based pension
allocation, the findings highlight social mobility as a key source of
legitimation for commodified pensions. Second, the article advances
research on market justice and market-based welfare preferences by
moving beyond static class locations and showing how intergenerational
occupational mobility structures support for pension commodification in
ways that partly echo, but also refine, existing evidence on class,
income and financial returns. Third, the null results for the moderating
role of meritocratic perceptions nuance accounts that tightly link
mobility, meritocracy and the embrace of neoliberal welfare: in this
context, mobility effects on pension commodification preferences appear
to operate independently of meritocratic beliefs. Finally, the study
contributes methodologically by applying Breen and Ermisch's
(\citeproc{ref-breen_effects_2024}{2024}) causal framework of mobility
effects to survey data, formalising mobility as origin-specific
treatment effects and estimating these using linear probability models
with inverse probability weighting.

Despite these contributions, the study has limitations that should be
acknowledged and read as avenues for future improvement. First, the
causal estimates hinge on the construction of inverse probability
weighting within Breen and Ermisch's
(\citeproc{ref-breen_effects_2024}{2024}) framework. Identification
relies on conditional independence given a relatively limited set of
origin-side covariates, together with positivity and SUTVA. Entropy
balancing helps to enforce covariate balance and mitigate selection on
observables, but unmeasured confounding cannot be ruled out. Sensitivity
analyses suggest moderate robustness to omitted variables; therefore,
the estimates should be interpreted with caution. Second, while the
analysis identifies mobility effects on preferences, it does not explain
why or how they arise. The results are analysed through material
self-interest and acculturation mechanisms. Still, the study does not
implement formal causal mediation analysis, nor can it determine whether
meritocratic perceptions function as mediators rather than mere
correlates. Third, the findings are context-specific: they are derived
from Chile, a highly unequal and strongly marketised pension regime,
which enhances internal validity but limits external validity. The
patterns documented here, therefore, cannot be straightforwardly
generalised to countries with different welfare institutions, mobility
regimes or levels of inequality, and should be seen as a theoretically
informed case study rather than a cross-national pattern.

Future research could extend these findings in several directions.
First, there is a need for richer measures of preferences for pension
commodification. This study relies on a single item that conflates
income-based allocation and fairness judgments; better items that
separate justice principles and dimensions of commodification would
allow a finer assessment of how people evaluate pensions as
income-conditioned commodities. Second, methodological limitations could
be addressed by including more detailed origin-side covariates and
life-course data, repeated measures of meritocratic perceptions, and
causal mediation analyses that unpack mechanisms such as material
self-interest, risk exposure, and meritocratic frames, alongside
alternative measures of mobility based on class schemas, income
trajectories, or wealth. Third, comparative designs would clarify how
institutional context shapes the attitudinal consequences of
intergenerational mobility. Applying similar causal strategies in
countries with different welfare regimes, levels of inequality, mobility
patterns, and degrees of pension marketisation would show whether the
Chilean pattern is context-specific or part of a broader configuration
linking mobility, merit and support for market-based welfare.
Ultimately, clarifying when and for whom market allocation of pensions
is judged fair is crucial to understanding the legitimate reach of
welfare commodification in unequal, marketised societies.

@@

>>>>

  \author{Andreas Laffert
Tamayo\thanks{This M.A. thesis in Sociology is one of my first achievements on the long path toward becoming an academic committed to the study of inequalities and to a public sociology. This journey would not have been possible without my relentless effort and the support of my family, friends and colleagues. Last but certainly not least, I want to acknowledge the motivation behind my research: the everyday processes through which neoliberalism is legitimised by ordinary people. This project was made possible through the financial support of ANID (BECAS/MAGÍSTER NACIONAL 2024) and the intellectual support and academic environment provided by the FONDECYT Regular project Nº 1250518. Any errors or omissions in this work are my sole responsibility. Comments are welcome at: andreas.laffert@uchile.cl}}
            \affil{%
                  Instituto de Sociología, Pontificia Universidad
                  Católica de Chile
              }

\date{December, 2025}

\usepackage{xcolor}
\definecolor{mylinkcolor}{HTML}{413395}
\hypersetup{
  colorlinks = true,
  linkcolor  = mylinkcolor,
  citecolor  = mylinkcolor,
  urlcolor   = mylinkcolor
}
\begin{document}
\maketitle
\begin{abstract}
The privatisation and marketisation of social services have been linked
to feedback effects on public attitudes towards welfare and market
inequality. From a moral-economy perspective, prior research shows that
support for market-based access to core services depends on
socio-structural position, inequality beliefs and institutional context.
Yet, we know little about how citizens judge the fairness of pension
commodification, and in particular how intergenerational mobility shapes
these judgements and whether meritocratic perceptions condition this
relationship. I address this gap by analysing preferences for
market-based pension access in Chile, a highly unequal country with
constrained mobility and a private pension system based on mandatory
individual capitalisation that depends heavily on contribution
histories. Drawing on three waves of the Chilean Longitudinal Social
Survey (ELSOC; 3,435 observations on 1,787 individuals) and under
explicit identification assumptions, I apply a recent causal framework
for mobility effects that treats destination class as a treatment
conditional on origin class, and estimate heterogeneous destination
effects on support for pension commodification. Results show that upward
mobility from middle to upper occupational status increases support for
market-based pension access relative to immobility. In contrast,
downward mobility from upper to middle or lower status reduces it.
Meritocratic perceptions do not significantly moderate these effects,
suggesting that mobility and meritocracy operate through largely
separate pathways. These findings advance research on welfare
marketisation by providing causal evidence on how intergenerational
mobility shapes preferences for market-based welfare, and by assessing
the potential moderating role of meritocratic perceptions in a highly
commodified regime. \newline \textbf{Keywords}: Pension commodification
· Intergenerational mobility · Causal inference · Meritocracy · Chile
\end{abstract}

\setstretch{1.5}
\section{Introduction}\label{introduction}

@@

Social mobility and meritocratic perceptions are closely intertwined in
the acculturation mechanism, although the direction of this link can
vary across trajectories and contexts. A large body of work consistently
shows that upwardly mobile individuals tend to interpret their
trajectory through a meritocratic lens, reading their success as
evidence that effort and ability are rewarded and reinforcing
individualistic attributions of poverty and wealth
(\citeproc{ref-bucca_merit_2016}{Bucca, 2016};
\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b};
\citeproc{ref-mijs_belief_2022}{Mijs et al., 2022};
\citeproc{ref-traini_climbing_2025}{Traini et al., 2025}), even in a
highly unequal and low mobility setting were the narratives of
individual success reinforces (\citeproc{ref-deng_its_2025}{Deng \&
Wang, 2025}). Studies of economic elites in Chile similarly report
strong endorsement of meritocracy, with respondents attributing their
ascent mainly to talent and business or leadership skills
(\citeproc{ref-atria_economic_2020}{Atria et al., 2020}). Evidence on
downward or blocked mobility is more ambivalent. Exposure to
low-mobility contexts and experiences of downward mobility are
associated with weaker beliefs that effort is rewarded, lower just-world
and opportunity beliefs, and stronger structural attributions for
success and failure (\citeproc{ref-davidai_why_2018}{Davidai, 2018};
\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b};
\citeproc{ref-mijs_how_2022}{Mijs \& Hoy, 2022}). Yet, under strong
meritocratic convictions, status loss can also be internalised as
personal failure, sustaining system-justifying attitudes and the
legitimation of existing hierarchies as a psychological defence strategy
(\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
\citeproc{ref-mau_inequality_2015}{Mau, 2015};
\citeproc{ref-traini_climbing_2025}{Traini et al., 2025}). In sum,
upward mobility is generally associated with stronger meritocratic
perceptions, whereas downward mobility is more closely linked to
structural explanations of inequality, without necessarily eroding
meritocratic convictions.

The ways in which social mobility trajectories shape interpretations of
meritocracy can determine whether mobility translates into stronger or
weaker support for market-based pension welfare. Following Mau's
(\citeproc{ref-mau_inequality_2015}{2015}) account of the ``majority
class'', upward mobility is expected to strengthen market-based welfare
not only because material interests change, but because those who rise
come to read their story as one of individual effort and talent
rewarded. In this view, markets are seen as fair arenas that properly
allocate pensions and other welfare benefits according to merit, so
upwardly mobile individuals with strong meritocratic convictions should
be especially prone to endorse commodified pensions. In contrast,
downward mobility typically fosters de-commodifying demands, yet under
strong meritocratic perceptions, status loss may be read as individually
deserved, weakening---or even reversing---its usual de-commodifying
pattern, as market outcomes are still viewed as broadly fair regardless
of direction. Empirical research on redistribution and attitudes toward
inequality supports this mechanism. Survey studies suggest that
subjective social mobility shapes support for redistribution and pension
spending through self-serving bias
(\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011};
\citeproc{ref-traini_climbing_2025}{Traini et al., 2025}), although
recent work indicates that this mechanism is evident only under explicit
constraints in experimental designs
(\citeproc{ref-molina_its_2019}{Molina et al., 2019}). More in line with
the acculturation theories, experimental reserach has shown that
exposure to upward-mobility information or self-made narratives reduces
support for redistribution, increases tolerance of inequality and system
justification, and even legitimises exploitative work conditions,
whereas exposure to low or downward mobility produces the opposite
pattern; crucially, these effects are partly mediated by stronger
meritocratic perceptions (\citeproc{ref-deng_its_2025}{Deng \& Wang,
2025}; \citeproc{ref-matamoros-lima_social_2025}{Matamoros-Lima et al.,
2025}; \citeproc{ref-shariff_income_2016}{Shariff et al., 2016}).
However, recent research has tested Piketty's (1995) hypothesis that
meritocracy would mediate the effect of upward social mobility on
redistributive preferences, without finding support for it
(\citeproc{ref-rodriguez_intergenerational_2024}{Rodriguez \&
Matilla-Garcia, 2024}). In this framework, meritocratic perceptions
condition the attitudinal consequences of mobility by amplifying the
pro-market effects of upward mobility and muting the critical potential
of downward mobility; thus, I treat meritocratic perceptions as a
moderating mechanism linking intergenerational mobility to preferences
for pension commodification.

@@

The findings reveal a tension in Chileans' judgements of market-based
pension welfare. Across 2016, 2018 and 2023, large majorities reject the
claim that it is fair for higher-income people to obtain better pensions
than those with lower incomes. However, a non-trivial---and
growing---minority endorses this principle, with the sharpest increase
between 2018 and 2023. Market-based pension allocation thus remains a
minority view. However, it is becoming more common, in line with
longitudinal evidence of rising support for income-based access to
welfare services in Chile
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}). This coexistence unfolds within a system of mandatory, privately
managed, defined-contribution accounts that tightly link old-age income
to earnings, contribution density and financial returns
(\citeproc{ref-galvez_pensiones_2024}{Gálvez et al., 2024};
\citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al., 2019}),
and that frame pensions as personal investments rather than
solidarity-based rights, promoting individual responsibility,
self-economisation, contribution-based desert and casting workers as
employee-entrepreneurs who must navigate financial risks and ``make
their money work'' (\citeproc{ref-borzutzky_pension_2012}{Borzutzky,
2012}). The Chilean case thus fits evidence that income-based welfare
access is more strongly endorsed in liberal, marketised regimes
(\citeproc{ref-lindh_public_2015}{Lindh, 2015}).This coexistence also
echoes recent pension conflict---from the ``No+AFP'' protests to
post-pandemic withdrawals and the ``Con mi plata NO'' mobilisation
(\citeproc{ref-galvez_con_2023}{Gálvez, 2023})---and illustrates mixed
policy feedback: the same regime generates rights-based critiques and
proprietarian defences of pension commodification. Together, these
patterns suggest that, despite discontent and a legitimacy crisis,
support for pension commodification remains a salient and contested
orientation, raising the question of which social trajectories and
beliefs underpin it.

@@

The second set of hypotheses examined whether meritocratic perceptions
condition the effects of intergenerational mobility. Building on
acculturation accounts and self-serving bias arguments
(\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}), I expected mobility to affect preferences for
pension commodification not only through material self-interest but also
through shifts in normative frames of distributive justice. Prior
research shows that upwardly mobile individuals tend to interpret their
trajectory through a meritocratic lens, reading their success as
evidence that effort and ability are rewarded and reinforcing
individualistic attributions of poverty and wealth. In contrast,
downward or blocked mobility is more often associated with structural
explanations of inequality and more ambivalent meritocratic beliefs,
sometimes including the internalisation of status loss as personal
failure (\citeproc{ref-davidai_why_2018}{Davidai, 2018};
\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
\citeproc{ref-deng_its_2025}{Deng \& Wang, 2025};
\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b};
\citeproc{ref-mijs_how_2022}{Mijs \& Hoy, 2022};
\citeproc{ref-traini_climbing_2025}{Traini et al., 2025}). Following
Mau's (\citeproc{ref-mau_inequality_2015}{2015}) account of the
``majority class'', I therefore expected upward mobility under strong
meritocratic convictions to be especially conducive to endorsing
commodified pensions, and strong meritocracy to dampen the usual
de-commodifying effect of downward mobility. Empirically, however, the
interaction models offer little support for this mechanism. Meritocratic
perceptions do not systematically moderate the mobility effects
identified in the primary analysis, and the few trajectory-specific
patterns that tentatively point in the expected direction are small and
imprecisely estimated. This result contrasts with survey and
experimental studies showing that upward mobility experiences or
exposure to self-made narratives reduce support for redistribution and
public pension spending partly by reinforcing meritocratic convictions
(\citeproc{ref-deng_its_2025}{Deng \& Wang, 2025};
\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
\citeproc{ref-matamoros-lima_social_2025}{Matamoros-Lima et al., 2025};
\citeproc{ref-molina_its_2019}{Molina et al., 2019};
\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011};
\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}), and aligns
with recent research that has shown that meritocracy does not mediate
the effect of social mobility on redistributive preferences
(\citeproc{ref-rodriguez_intergenerational_2024}{Rodriguez \&
Matilla-Garcia, 2024}). One possibility is that, in this context,
intergenerational occupational mobility---as operationalised here---has
limited impact on meritocratic perceptions, which appear more weakly
tied to mobility trajectories than previous research suggests and more
strongly anchored in broader socialisation processes and class-specific
cultures.

The contributions of this study can be summarized as follows.
Substantively, it offers one of the first systematic analyses of
preferences for pension commodification in Chile's highly unequal and
marketised pension regime, addressing Mau's
(\citeproc{ref-mau_inequality_2015}{2015}) question of why core elements
of neoliberal welfare arrangements gain public acceptance. By showing
that upward and downward intergenerational mobility into and out of
upper-class positions causally shape support for market-based pension
allocation, the findings highlight social mobility as a key source of
legitimation for commodified pensions. Second, the article advances
research on market justice and market-based welfare preferences by
moving beyond static class locations and showing how intergenerational
occupational mobility structures support for pension commodification in
ways that partly echo, but also refine, existing evidence on class,
income and financial returns. Third, the null results for the moderating
role of meritocratic perceptions nuance accounts that tightly link
mobility, meritocracy and the embrace of neoliberal welfare: in this
context, mobility effects on pension commodification preferences appear
to operate independently of meritocratic beliefs. Finally, the study
contributes methodologically by applying Breen and Ermisch's
(\citeproc{ref-breen_effects_2024}{2024}) causal framework of mobility
effects to survey data, formalising mobility as origin-specific
treatment effects and estimating these using linear probability models
with inverse probability weighting.

Despite its merits, the study has limitations that should be
acknowledged and read as avenues for future improvement. First, the
causal estimates hinge on the construction of inverse probability
weighting within Breen and Ermisch's
(\citeproc{ref-breen_effects_2024}{2024}) framework. Identification
relies on conditional independence given a relatively limited but
relevant set of origin-side covariates, together with positivity and
SUTVA. Entropy balancing helps to enforce covariate balance and mitigate
selection on observables, but unmeasured confounding cannot be ruled
out. Sensitivity analyses suggest moderate robustness to omitted
variables; therefore, the estimates should be interpreted with caution.
Second, while the analysis identifies mobility effects on preferences,
it does not explain why or how they arise. The results are analysed
through material self-interest and acculturation mechanisms. Still, the
study does not implement formal causal mediation analysis, nor can it
determine whether meritocratic perceptions function as mediators rather
than mere correlates. Third, the findings are context-specific: they are
derived from Chile, a highly unequal country with a strongly marketised
pension regime, which enhances internal validity but limits external
validity. The patterns documented here, therefore, cannot be
straightforwardly generalised to countries with different welfare
institutions, mobility regimes or levels of inequality, and should be
seen as a theoretically informed case study rather than a cross-national
pattern.

Future research could extend these findings in several directions.
First, there is a need for richer measures of preferences for pension
commodification. This study relies on a single item that conflates
income-based allocation and fairness judgments; better items that
separate justice principles and dimensions of commodification would
allow a finer assessment of how people evaluate pensions as
income-conditioned commodities. Second, methodological limitations could
be addressed by including more detailed origin-side covariates and
life-course data, repeated measures of meritocratic perceptions, and
causal mediation analyses that unpack mechanisms such as material
self-interest, risk exposure, and meritocratic frames, alongside
alternative mobility measures based on class schemas, income
trajectories, or wealth. Third, comparative designs would clarify how
institutional context shapes the attitudinal consequences of
intergenerational mobility. Applying similar causal strategies in
countries with different welfare regimes, levels of inequality, mobility
patterns, and degrees of pension marketisation would show whether the
Chilean pattern is context-specific or part of a broader configuration
linking mobility, merit and support for market-based welfare
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Ultimately, clarifying
when and for whom market allocation of pensions is judged fair is
crucial to understanding the legitimate reach of welfare commodification
in unequal, marketised societies.

@@

\bibitem[\citeproctext]{ref-traini_climbing_2025}
Traini, C., Ehmes, S., \& Gangl, M. (2025). Climbing the ladder or
falling behind: {How} social mobility shapes perceptions of meritocracy
in the wake of rising inequality. \emph{PO- LAR Working Paper \#7.
Frankfurt: Goethe University. Retrieved from Www.polar-Project.org,
Version Dated 12 September 2025}.
